---
title: "Introduction to Relational Event Models"
author: "Melania Lembo, Martina Boschi, Ernst C. Wit"
date: "2026-06-22"
output:
  md_document:
    variant: markdown_strict
    preserve_yaml: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(output = function(x, options) {
  paste0('<div class="r-output"><pre><code>', x, '</code></pre></div>')
})
```

## Introduction

In this practical, we will show how to use relational event models beyond linear effect specification, in the context of the first time invasion of species into regions over time. (Boschi et al., 2025). We will show how to fit models with time-varying, non-linear and random effects to gain insight on how climatological, economical and geographical aspect might explain the dynamics of this ecological process.

---

## 1. Package requirements and setup

If you are running this tutorial locally, make sure to uncomment the `remotes::install_github command` below to install the package.

```{r warning = FALSE}
# install.packages("remotes")
# install.packages("pbapply")
# remotes::install_github("franciscorichter/amorem")

library(amorem)
library(pbapply)
```

---

## 2. Data uploading, inspection and cleaning

Load the input data `input00.RData` located in the `01-Data/01-Inputs` folder. Inspect the original data contained in . Have a look at the following columns of the object `FR`: `LifeForm`, `Taxon`, `Region`, `FirstRecord`, `Source`.

```{r}
load(file="01-Data/01-Inputs/input00.RData")

head(FR[,c("LifeForm", "Taxon",
           "Region", "FirstRecord",
           "Source")])
```

We firstly show how data arrives from the original source (Seebens et al., 2018). These data involve 17 life forms. For each record, we are able to identify the **species** (`Taxon`) of a certain **lifeform** (`Lifeform`) that is invading a **region** (`Region`) in which it was not native and was not seen before the reported **year** (`FirstRecord`). The Alien Species First Records Database collects several sources: the source of each FR is reported in the column `Source`.

A **first record** (FR) is the year $t$ when a (non-native) species $s$ first invades region $r$. The following dataset, `FR`, already contains information we need to interpret first records as *relational events* $(s,r,t)$.

Particularly, the *sender set* $V^S$ is composed of the unique recorded species, while regions of the world (countries and islands) form the *receiver set* $V^R$. The *time window* spans from $t_0=1880$ to $T=2005$.

A species' **native range** (NR) is the collection of areas where it is indigenous. Slightly more liberally, we refer to native range as the set of sites where a species was already present before the start of the analysis period, which in this context is $1880$. The object `native` contains such information. 

```{r}
native[sample(1:nrow(native), 10), c("species", "region")]
```

This information is crucial for the definition of the *risk set*. If a species is native in a region, then such dyad is never at risk of occurring ever during our analyzed time window. Additionally, the relational event is **non-recurrent**, namely, the event first record, for its definition, occurs only once.
This also means that the risk set is reducing over time: once the event is observed, it cannot be sampled anymore as a non-event. 
Also, if interested in computing *endogenous covariates* (that depend on prior interactions), such information is necessary for the events occurring at the beginning (1880).

In this tutorial, in order to make the analysis easy to reproduce, we focus on **insects** only. `first_records` contains the insect FRs of interest.

```{r}
head(first_records[,c("year","lf",
           "species", 
           "region")])
```

We know format the data in a convenient manner for analysis using `amorem` package. We need to make sure that both the data structures for the sequence of relational events and the set of dyads not at risk at the beginning of the time-window contain a `sender` and `receiver` column. 

```{r}
event_log <- standardize_event_log(first_records[,c(1,3:4)],
                                   sender_col = "species",
                                   receiver_col = "region",
                                   time_col = "year")
  
head(event_log)
```

```{r}
native_df<-native[, c("species", "region")]
colnames(native_df)<-c("sender", "receiver")
head(native_df)
```

In addition we have information on 2 dyadic attribute, **distance** and **trade** between regions, and on a species-level attribute, **temperature**.

Regarding climate, `data_temperature` contains the names of the countries in the first column and the average values of near-surface air temperature in the second column (Watanabe et al., 2011).

```{r}
head(data_temperature)
```

For trade, the source information, derived from Barbieri et al. (2009), `data_trade` reports trade flows among countries.

```{r}
head(data_trade)
```

Distance between regions is computed referring to the closest borders and consequently is equal to zero for neighbours. Source data for this driver consists of the `R` package `geosphere` (Hijmans et al., 2017).`data_distance` is a matrix with dimension names referring to the regions. The corresponding element in the matrix reports the distance between the two.

```{r}
data_distance[1:3,1:3]
```

We will leverage this information to investigate the effect of temperature, trade and distance in the dynamics of species invasion.

---

## 3. Case-control sampling from the risk set

To perform inference by maximizing the *sampled partial likelihood* through the fitting of a *degenerate logistic* model we need to perform *case-1-control sampling*. In particular, for each observed event, we sample uniformly at random from the risk set at the time of the event, a *non-event* dyad.

Given the time granularity, we have that multiple invasions occur the same year. In order to address these *ties*, for each event, we exclude from the risk set at that time, all the first records that occurred in the same year.

```{r}
set.seed(1234)
cc <- sample_non_events(event_log,
                        n_controls = 1,              # 1 control is default
                        scope         = "all",
                        mode          = "two",
                        risk          = "remove",          # drops past + concurrent
                        exclude_pairs = native_df[,1:2])   ### excludes native dyads


head(cc)
```

Notice the long format of the case-control dataset: events and non-events appear on different rows. In particular each event row (`event = 1`) is followed by the corresponding sample non-event (`event = 0`).

---

## 4.  Computing endogenous covariates leveraging exogenous information

We now will combine the exogenous attributes with the relational event info and define 3 endogenous covariates, describing aspects of trade, distance and temperature that might explain invasion dynamics.

**Climate dissimilarity (via temperature).** This covariate, given a relational event (and all the relevant information), computes the minimum temperature difference between the region involved and the previously invaded regions by the considered species. It depends on both the sender and the receiver of the event and can therefore be considered a *dyad-level covariate*. It is *time-dependent* but does not incorporate information about internal time. It is *endogenous*, as it relies on the regions previously invaded by the species, but it also incorporates exogenous information.

**Trade.** In the existing literature, **international trade** has been recognized as a key factor in explaining the spread of alien species. Here, we consider trade as the yearly commerce between already-invaded territories and the involved region. This also is a dyad-level, endogenous covariate.

**Distance.** In this instance, given a relational event (and all the relevant information that is required), we compute the shortest distance among the regions in which the species is already present at that time and log-transform it.

Notice that the endogenous component is common to all of them and it involves **past invaded regions by the considered species**. For the invasions occurring in 1880, the past invaded regions are the ones included in `native` for that species. To make sure the function treats it as such we hard-code a `time` column for these dyads with value 1879.
In addition, note that such covariate computation is to be done for both events and non-events. For the latter ones, the past information must always be only relative the actual observed events.

```{r}
## Invaded regions ####
native_df<-data.frame(native_df, time = "1879")
feats <- compute_endogenous_features(cc, stats = "sender_receivers_set",
                                         history_log = event_log, prior_log = native_df)

## remove current region to avoid zero difference
feats$invaded<- mapply(setdiff,feats$sender_receivers_set,feats$receiver)

print(feats[1:3,])

```
The following code computes, for each event/non-event, the minimum absolute difference between the current region and all past invaded regions by the considered species. We refer to it as **climate dissimilarity**.
```{r}
## Temperature ####

compound_region_map <- list(
  "USACanada" = c("United States", "Canada")
)

get_temp <- function(region,compound_areas = compound_region_map) {
  if (region %in% data_temperature$X) {
    return(data_temperature[data_temperature$X == region, "temp"][1])
  }
  if (region %in% names(compound_areas)) {
    parts <- compound_region_map[[region]]
    temps <- data_temperature[data_temperature$X %in% parts, "temp"]
    if (length(temps) > 0) return(mean(temps, na.rm = TRUE))
  }
  return(NA_real_)
}

dt_value <- mapply(function(invaded_regions, current_region) {
  
  if (length(invaded_regions) == 0) return(NA_real_)
  
  avg_temp_invaded  <- sapply(invaded_regions, get_temp)
  avg_temp_interest <- get_temp(current_region)
  
  if (is.na(avg_temp_interest) || all(is.na(avg_temp_invaded))) 
    return(NA_real_)
  
  min(abs(avg_temp_invaded - avg_temp_interest), na.rm = TRUE)
  
}, feats$invaded, feats$receiver)

feats$temp  <- dt_value
```

The following code computes the **trade** variable that captures the log-transformed sum of the most recent recorded trade flows into a region from each previously invaded region.
A few notes on data pre-processing:

- Negative values of trade are considered equal to $0$.

- When a pair of nations is never mentioned in the source data, the trade volume is presumed to be $0$, based on the idea that if a value is never recorded, it should be negligible.

- When the pair is mentioned but there is a non-imputed gap at the time of interest, the trade flow is evaluated based on the latest available year for which information is available.

- Because trade varies by orders of magnitude, we take a log-transformation for the analysis.

```{r}
## Trade ####

t <- which(data_trade$transfer < 0)
data_trade$transfer[t] <- 0
trade_value <- pbmapply(function(invaded_regions, current_region, y) {
  trade_funct_new <- function(to_region) {
    if (length(invaded_regions) == 0) return(0)
    x <- data_trade[
      data_trade$FromRegion %in% invaded_regions &
        data_trade$ToRegion   == to_region &
        data_trade$year       <= y,
    ]
    if (nrow(x) == 0) return(0)
    most_recent <- aggregate(x$year, list(x$FromRegion), FUN = max)
    trade_vals <- mapply(function(region, max_year) {
      x$transfer[x$FromRegion == region & x$year == max_year]
    }, most_recent[, 1], most_recent[, 2])
    log(sum(unlist(trade_vals), na.rm = TRUE) + 1)
  }
  if (current_region == "USACanada") {
    mean(c(trade_funct_new("United States"),
           trade_funct_new("Canada")))
  } else {
    trade_funct_new(current_region)
  }
}, feats$invaded, feats$receiver, feats$time)

feats$trade<- trade_value
```

Now we account for the geographic aspect, through **distance**, which is compute as the logged minimum geographic distance between the current region and any previously invaded region. 

```{r}
## Distance ####

distance_value <- pbmapply(function(invaded_regions, current_region, y) {
  
  if (length(invaded_regions) == 0) return(NA_real_)
  
  distances <- data_distance[current_region, invaded_regions]
  
  log(min(distances, na.rm = TRUE) + 1)
  
}, feats$invaded, feats$receiver, feats$time)

feats$dist <- distance_value

```

For fitting the models, it is convenient to transform the case-1-control dataset and related covariates, contained in `feats` from long-format (event on one row and corresponding non-event in the following row) to wide-format, in which the event and non-event appear in the same row.

```{r}
cc_wide <- widen_case_control(feats, case = "event",stratum = "stratum")
head(cc_wide)
```

## 4.  Fitting models with flexible effects using degenerate logistic trick

We will now fit 3 separate models, each of them with one of the 3 covariates and specify different type of effects.

Let us start with a model that assume a **linear effect on climatic dissimilarity**:

```{r}
## Fixed linear effect of Temperature ####
m1_only_temp_l<- rem(~ temp, data = cc_wide, method = "gam")
summary(m1_only_temp_l)
```

The coefficient for climatic dissimilarity, which remains constant over time, is negative. This indicates that the rate of invasions tends to increase if the species has already invaded countries with similar temperatures.

Now consider a model with a **time-varying effect for trade**. Note the effect changes over time, potentially non-linearly, but it is linear in the covariate. `tv()` and the specification of `time=...` are required for such an effect, when using `rem()` function.

```{r tve_trade}
## Time-varying linear effect ####
m2_only_trade_tv<- rem(~ tv(trade),time = "time_ev", data = cc_wide, method = "gam")
summary(m2_only_trade_tv)
plot(m2_only_trade_tv)
```

The effect of trade is positive throughout but, appears to be decreasing over the time period. So greater trade flows between regions, increase the chance of an invasion but such trend used to be strongeri in the past. It could be the result of introduction of international norms and laws that regulate non-indigenous species importation.

We now fit a model with a **non-linear effect of distance**. Notice the `nl()` to specify it.

```{r nl_dist}
## Non-linear effect of distance ####
m3_only_dist_nl<- rem(~ nl(dist), data = cc_wide, method = "gam")
summary(m3_only_dist_nl)
plot(m3_only_dist_nl)
```

From the estimated non-linear effect, which presents an overall descreasing trend, we have that the greater the distance between 2 countries the less likely it is for species to invade from one to other.

We now consider a model with a **sender-level random intercept**. The purpose of random effects is to account for node/dyad-level heterogeneity. In this case, the assumption is that not all species are the same, in the sense that some might have a tendency to invade more than others. Such a random intercept accounts for that and is specified in the code using `re()`:

```{r}
m4_only_sender_re<- rem(~ re(sender), data = cc_wide, method = "gam")
summary(m4_only_sender_re)
re.species <- coefficients(m4_only_sender_re)

#### rem with re() effect specification fits a random intercept based on the specified group var
#### it handles factor encoding internally which renders used of output less intuitive.
#### the following replicates the mapping from group var format to factor in order to print results in the original format
ev <- cc_wide[["sender_ev"]]
nv <- cc_wide[["sender_nv"]]
fmat <- factor(c(as.character(ev), as.character(nv))) # Replicate what rem() does internally
names(re.species) <- levels(fmat)  # The mapping: index 1 = levels(fmat)[1], index 2 = levels(fmat)[2], etc.
```

The 5 most and less invasive species:
```{r}
print(sort(re.species, decreasing = TRUE)[1:5])
print(sort(re.species)[1:5])
```
We conclude this section by fitting a model with all 4 effects combined and perform model comparison using AIC.

```{r}
## Full model ####
m5_full <- rem(~ temp + 
                 tv(trade) + 
                 nl(dist) +
                 re(sender), time = "time_ev", data = cc_wide, method = "gam")

aic_table <- data.frame(
  model     = c("dt_only", "tr_only", "d_only", "sp_only", "complete"),
  AIC       = c(AIC(m1_only_temp_l), AIC(m2_only_trade_tv), AIC(m3_only_dist_nl),
                AIC(m4_only_sender_re), AIC(m5_full))
)
aic_table <- aic_table[order(aic_table$AIC), ]
print(aic_table)
```

Among the 5 fitted models, the complete one has better predictive performance.

