---
title: 'Simulation Study: Implementation'
author: "Boschi, Martina"
date: "2026-04-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synthetic Data: Academics moving in US states

The following simulation study includes: 
-   First part: REM including a linear effect for reciprocity and a non-linear effect for distance;
-   Second part: REM including a linear effect for reciprocity and a non-linear effect for distance plus a global effect for the day of the week. 

## Preliminarities

```{r pressure, echo=FALSE, message=FALSE, warning=FALSE}
source("script/packages.R")
```

### Members, Size, and Order of the temporal network

```{r message=FALSE, warning=FALSE}
source("script/states.R")
```

### True non-linear effect of distance

Distance is an example of exogenous information that is external to the relational event network. We exploit this information to compute an endogenous covariate. 

Source for distance: [Walker, 2024] [Pebesma, 2018] [Hijmans, 2022]

We assume the contribution of distance being a non-linear function of distance itself. 

The true value for the coefficients is retrieved in such a way the final contribution is similar to a negative contribution of the logarithm of the original covariate, namely:

```{r}
source("script/input_distance.R")
```

### Linear effect of weekday and reciprocity

The baseline of a interstate academic conference is assumed to be smaller during the weekends with respect to the weekdays.

```{r}
source("script/true_coefficients.R")
```

## Event Simulation (no shifting)

```{r}
event_simulator <- function(states, p, n, b0, b1, fit, seed) {
  
  set.seed(seed)
  print(seed)
  
  # some constants
  current.tm<-0
  rec<-matrix(0,ncol=p,nrow=p)
  
  #where to save
  dat<-NULL
  nondat<-NULL
  tms<-NULL
  tm = 0  # Initialize time to 0
  
  # Initialize the invaded list with a random sample of states
  last_contacted <- matrix(NA, nrow=n+1, ncol=p)
  last_contacted[1,] <- sample(states, p)
  colnames(last_contacted) <- states
  
  # Compute the distance matrix for each state and potential state pairs
  dist.m <- t(sapply(states, function(st) 
    sapply(states, function(pot.st)
      dist_matrix[pot.st, last_contacted[1,st]])))
  distance <- matrix(spline_function(as.vector(dist.m),
                                     fit), 
                     nrow=p, ncol=p)
  
  #simulation of events
  i<-1
  while (i <=n){
    
    if(i%%100==0) print(i)
    current.weekday<-1*(current.tm%%7>=2)
    hazard<- exp(b0*current.weekday+b1*rec+distance)/(p^2)
    
    # time increase
    dt<-rexp(1,sum(hazard))
    
    # check if this change meant that the event 
    # moved from weekend to weekday or vice-versa:
    new.weekday<-1*((current.tm+dt)%%7>=2)
    if (dt>5| new.weekday!=current.weekday){
      current.tm<- (current.tm%/%7)*7+2+current.weekday*5
    } else {
      current.tm<-current.tm+dt
      tms<-c(tms,current.tm)
      event.id<-sample(1:p^2, 1,prob = hazard/sum(hazard))
      s<-(event.id-1)%%p+1
      r<-(event.id-1)%/%p+1
      dat<-rbind(dat,c(current.tm,s,r,event.id,
                       current.weekday,
                       dist.m[s,r],
                       rec[s,r]))
      
      non.event.id <- sample(setdiff(1:p^2, event.id),1)
      non.s <- (non.event.id-1)%%p+1
      non.r <- (non.event.id-1)%/%p+1
      nondat <- rbind(nondat,c(non.s,non.r,non.event.id,
                               dist.m[non.s,non.r],
                               rec[non.s,non.r]))
      
      rec[r,s] <- rec[r,s] + 1
      last_contacted[i+1,] <- last_contacted[i,]
      last_contacted[i+1,states[s]] <- states[r]
      
      dist.m <- t(sapply(states, function(st) 
        sapply(states, function(pot.st)
          dist_matrix[pot.st, last_contacted[i+1,st]])))
      distance <- matrix(spline_function(as.vector(dist.m),
                                         fit), 
                         nrow=p, ncol=p)
      hazard<- exp(b0*current.weekday+b1*rec+distance)/(p^2)
      
      i<-i+1
    }
  }
  
  remdat <- cbind(1:n, dat, nondat,
                  dat[,6]-nondat[,4],
                  dat[,7]-nondat[,5],
                  1)
  colnames(remdat) <- c("index", "tms","s","r","event.id",
                        "weekday","distance", "rec",
                        "non.s","non.r","non.event.id",
                        "non.distance", "non.rec",
                        "delta.dist", "delta.rec","one")
  remdat <- as.data.frame(remdat)
  
  return(list(remdat, last_contacted))
}
```

```{r eval=FALSE}
n_sim = 100
set.seed(1234)
remdats <- sapply(1:n_sim, function(x)
  event_simulator(states = states, p = length(states), n = n,
                  b0 = b0, b1 = b1, fit, seed=x),
  simplify =FALSE)
save(remdats, file="output/remdats_data.RData")
```

## Model Fitting (no shifting)

For each of the simulated dataset, we fit a model including a linear effect fir reciprocity and a non-linear effect for distance. Particularly, for each event we did sample one non-event and the distance between the covariates is taken into account.

```{r eval=FALSE}
load(file="output/remdats_data.RData")
remdats_data <- lapply(remdats, function(x) x[[1]])
gam.fits <- lapply(remdats_data, function(x){
  dist_MAT <- cbind(x$distance, x$non.distance)
  ONE <- cbind(x$one, -x$one)
  fit <- gam(one ~ s(dist_MAT, by=ONE) + 
               delta.rec - 1,
             family="binomial",
             data=x)
  return(fit)
}
)
save(gam.fits, file="output/gam.fits.RData")
```

## Event Shifting

```{r}
remdat_shifting <- function(remdat, last_contacted, p, n, seed){
  
  set.seed(seed)
  
  T<-max(remdat$tms)
  shifts<-rexp(p^2,1/T)
  remdat_complete<-NULL
  
  for (i in 1:n){
    
    current.tm <- remdat[i,2]
    
    atrisk<-(shifts<current.tm+shifts[remdat[i,5]]) & 
            (shifts+T>current.tm+shifts[remdat[i,5]])
    
    non.event.id <- sample(1:p^2,1,prob = atrisk/sum(atrisk))
    non.s<-(non.event.id-1)%%p+1
    non.r<-(non.event.id-1)%/%p+1
    
    non.current.weekday<-1*((current.tm+shifts[remdat[i,5]]-
                               shifts[non.event.id])%%7>=2)
    non.rec<-sum((remdat[,2]<current.tm+shifts[remdat[i,5]]-
                    shifts[non.event.id])&(remdat[,3]==non.r)&
                   (remdat[,4]==non.s))
    
    dist.m <- t(sapply(states, function(st) 
      sapply(states, function(pot.st)
        dist_matrix[pot.st, last_contacted[i+1,st]])))
    
    remdat_complete<-rbind(remdat_complete,
                           unlist(c(remdat[i,1:8],
                                    non.s,non.r,non.event.id,
                                    non.current.weekday,
                                    dist.m[non.s,non.r],non.rec)))
  }
  
  colnames(remdat_complete) <- c(colnames(remdat)[1:8],
                                 "non.s","non.r","non.event.id",
                                 "non.weekday","non.distance","non.rec")
  remdat_complete <- as.data.frame(remdat_complete)
  
  remdat_complete$delta.dist = remdat_complete$distance - 
    remdat_complete$non.distance 
  remdat_complete$delta.wkday = remdat_complete$weekday - 
    remdat_complete$non.weekday
  remdat_complete$delta.rec = remdat_complete$rec - 
    remdat_complete$non.rec
  remdat_complete$one = 1
  
  return(remdat_complete) 
  
}
```

```{r eval=FALSE}
shifted_rems <- sapply(1:n_sim, function(x)
  remdat_shifting(remdats[[x]][[1]], remdats[[x]][[2]], p, n, seed=x), 
  simplify = FALSE)
save(shifted_rems, file="output/shifted_rems.RData")
length(shifted_rems)
```

## Model Fitting (shifting)

```{r eval=FALSE}
load(file="output/shifted_rems.RData")
gam.fits_shifted <- lapply(shifted_rems, function(x){
  dist_MAT <- cbind(as.numeric(x$distance), 
                    as.numeric(x$non.distance))
  ONE <- cbind(x$one, -x$one)
  fit <- gam(one ~ s(dist_MAT, by=ONE, k=11, bs="bs") + 
               delta.rec + delta.wkday - 1,
             family="binomial",
             data=x)
  return(fit)
}
)
save(gam.fits_shifted, file="output/gam.fits_shifted.RData")
```
