---
title: "Running Example: Data Generation"
author: "Boschi, M."
date: "2026-04-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synthetic Data: Academics moving in US states.

The following simulation study includes: 
-   First part: REM including a linear effect for reciprocity and a non-linear effect for distance;
-   Second part: REM including a linear effect for reciprocity and a non-linear effect for distance plus a global effect for the day of the week. 

The goal of this document is to implement several inference procedures, not just logistic regression. To achieve this, a dataset is constructed that includes, for each event, all possible non-events at risk at the same time. There is a column aimed at distinguishing among events and non-events. 

## Preliminarities

More details about the preliminaries can be found in the document: `01-simulation-study-implementation`.

```{r pressure, echo=FALSE, message=FALSE, warning=FALSE}
source("script/packages.R")
```

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
source("script/states.R")
source("script/input_distance.R")
source("script/true_coefficients.R")
```

```{r}
n_sim = 1
```

## Event Simulation (no shifting)

```{r}
event_simulator <- function(states, p, n, b0, b1, fit, seed, m,
                            cox=TRUE, clogit=TRUE) {
  
  set.seed(seed)
  print(seed)
  
  # some constants
  current.tm<-0
  rec<-matrix(0,ncol=p,nrow=p)
  
  #where to save
  dat<-NULL
  nondat<-NULL
  dat_cox <-NULL
  dat_sampled <- NULL
  dat_gam_multiple <- NULL
  tms<-NULL
  tm = 0  # Initialize time to 0
  
  # Initialize the invaded list with a random sample of states
  last_contacted <- matrix(NA, nrow=n+1, ncol=p)
  last_contacted[1,] <- sample(states, p)
  colnames(last_contacted) <- states
  
  # Compute the distance matrix for each state and potential state pairs
  dist.m <- t(sapply(states, function(st) 
    sapply(states, function(pot.st)
      dist_matrix[pot.st, last_contacted[1,st]])))
  distance <- matrix(spline_function(as.vector(dist.m),
                                     fit), 
                     nrow=p, ncol=p)
  
  #simulation of events
  i<-1
  while (i <= n){
    
    if(i%%10==0) print(i)
    current.weekday<-1*(current.tm%%7>=2)
    hazard<- exp(b0*current.weekday+b1*rec+distance)/(p^2)
    
    # time increase
    dt<-rexp(1,sum(hazard))
    
    # check if this change meant that the event 
    # moved from weekend to weekday or vice-versa:
    new.weekday<-1*((current.tm+dt)%%7>=2)
    if (dt>5| new.weekday!=current.weekday){
      current.tm<- (current.tm%/%7)*7+2+current.weekday*5
    } else {
      left <- current.tm
      
      current.tm<-current.tm+dt
      
      right <- current.tm
      
      tms<-c(tms,current.tm)
      event.id<-sample(1:p^2, 1,prob = hazard/sum(hazard))
      s<-(event.id-1)%%p+1
      r<-(event.id-1)%/%p+1
      dat<-rbind(dat,c(current.tm,s,r,event.id,
                       current.weekday,
                       dist.m[s,r],
                       rec[s,r]))
      
      non.event.id <- sample(setdiff(1:p^2, event.id),1)
      non.s <- (non.event.id-1)%%p+1
      non.r <- (non.event.id-1)%/%p+1
      nondat <- rbind(nondat,c(non.s,non.r,non.event.id,
                               dist.m[non.s,non.r],
                               rec[non.s,non.r]))
    
      eligible <- setdiff(1:p^2, event.id)
      
      if (cox==TRUE){
        dat_cox <- rbind(dat_cox, 
                       c(1,current.tm,s,r,event.id,
                       current.weekday,
                       dist.m[s,r],
                       rec[s,r],
                       left, right))
      for (non.event.id in eligible){
        non.s <- (non.event.id-1)%%p+1
        non.r <- (non.event.id-1)%/%p+1
        dat_cox <- rbind(dat_cox, 
                         c(0, current.tm,
                           non.s,non.r,non.event.id,
                           current.weekday,
                           dist.m[non.s,non.r],
                           rec[non.s,non.r], 
                           left, right))
      }
      }
      
      sampled_non_events <- sample(eligible, m)
      
      if(clogit==TRUE){
        dat_sampled <- rbind(dat_sampled, 
                       c(1, current.tm,s,r,event.id,
                       current.weekday,
                       dist.m[s,r],
                       rec[s,r],
                       left, right))
      
      for (non.event.id in sampled_non_events){
        non.s <- (non.event.id-1)%%p+1
        non.r <- (non.event.id-1)%/%p+1
        dat_sampled <- rbind(dat_sampled, 
                         c(0,current.tm,
                           non.s,non.r,non.event.id,
                           current.weekday,
                           dist.m[non.s,non.r],
                           rec[non.s,non.r],
                           left, right))
      }
      }
      
      rec[r,s] <- rec[r,s] + 1
      last_contacted[i+1,] <- last_contacted[i,]
      last_contacted[i+1,states[s]] <- states[r]
      
      dist.m <- t(sapply(states, function(st) 
        sapply(states, function(pot.st)
          dist_matrix[pot.st, last_contacted[i+1,st]])))
      distance <- matrix(spline_function(as.vector(dist.m),
                                         fit), 
                         nrow=p, ncol=p)
      hazard<- exp(b0*current.weekday+b1*rec+distance)/(p^2)
      
      i<-i+1
    }
  }
  
  remdat <- cbind(1:n, dat, nondat,
                  dat[,6]-nondat[,4],
                  dat[,7]-nondat[,5],
                  1)
  colnames(remdat) <- c("index", "tms","s","r","event.id",
                        "weekday","distance", "rec",
                        "non.s","non.r","non.event.id",
                        "non.distance", "non.rec",
                        "delta.dist", "delta.rec","one")
  remdat <- as.data.frame(remdat)
  
  if(cox==TRUE){
    colnames(dat_cox) <- c("event", "tms",
                           "s","r","event.id",
                           "weekday",
                           "distance",
                           "rec", 
                         "start", "stop")
    dat_cox <- as.data.frame(dat_cox)
  }
  
  if(clogit==TRUE){
    colnames(dat_sampled) <- c("event", "tms",
                           "s","r","event.id",
                           "weekday",
                           "distance",
                           "rec", 
                         "start", "stop")
    dat_sampled <- as.data.frame(dat_sampled)
  }
  
  return(list(remdat, last_contacted, 
              dat_cox, dat_sampled))
}
```

```{r eval=FALSE}
set.seed(1234)
remdats_cox <- sapply(1:n_sim, function(x)
  event_simulator(states = states, p = length(states), n = 100,
                  b0 = b0, b1 = b1, fit = fit, seed=10, m=5,
                  cox=TRUE, clogit=TRUE),
  simplify =FALSE)

remdats_nocox <- sapply(1:n_sim, function(x)
  event_simulator(states = states, p = length(states), n = n,
                  b0 = b0, b1 = b1, fit = fit, seed=10, m=5,
                  cox=FALSE, clogit=FALSE),
  simplify =FALSE)

save(remdats_cox, remdats_nocox, file="output/running_example_data.RData")
```

```{r eval=FALSE}
load(file="output/running_example_data.RData")
all(remdats_cox[[1]][[1]]$s == remdats_nocox[[1]][[1]]$s[1:100])
all(remdats_cox[[1]][[1]]$r == remdats_nocox[[1]][[1]]$r[1:100])
remdats_data <- lapply(remdats_nocox, function(x) x[[1]])
gam.fits <- lapply(remdats_data, function(x){
  dist_MAT <- cbind(x$distance, x$non.distance)
  ONE <- cbind(x$one, -x$one)
  fit <- gam(one ~ s(dist_MAT, by=ONE) + 
               delta.rec - 1,
             family="binomial",
             data=x)
  return(fit)
}
)
save(gam.fits, file="output/running_example_gam.fits.RData")
```

```{r}
remdat_shifting <- function(remdat, last_contacted, p, n, seed){
  
  set.seed(seed)
  
  T<-max(remdat$tms)
  shifts<-rexp(p^2,1/T)
  remdat_complete<-NULL
  
  for (i in 1:n){
    
    current.tm <- remdat[i,2]
    
    atrisk<-((shifts<current.tm+shifts[remdat[i,5]]) & 
             (shifts+T>current.tm+shifts[remdat[i,5]]))
    
    non.event.id <- sample(1:p^2,1,prob = atrisk/sum(atrisk))
    non.s<-(non.event.id-1)%%p+1
    non.r<-(non.event.id-1)%/%p+1
    
    non.current.weekday<-1*((current.tm+shifts[remdat[i,5]]-
                               shifts[non.event.id])%%7>=2)
    non.tms <- current.tm+shifts[remdat[i,5]]- shifts[non.event.id]
    non.rec<-sum((remdat[,2]<current.tm+shifts[remdat[i,5]]-
                    shifts[non.event.id])&(remdat[,3]==non.r)&
                   (remdat[,4]==non.s))
    
    dist.m <- t(sapply(states, function(st) 
      sapply(states, function(pot.st)
        dist_matrix[pot.st, last_contacted[i+1,st]])))
    
    remdat_complete<-rbind(remdat_complete,
                           unlist(c(remdat[i,1:8],
                                    non.tms, non.s,non.r,non.event.id,
                                    non.current.weekday,
                                    dist.m[non.s,non.r],non.rec)))
  }
  
  colnames(remdat_complete) <- c(colnames(remdat)[1:8],
                                 "non.tms", "non.s","non.r","non.event.id",
                                 "non.weekday","non.distance","non.rec")
  remdat_complete <- as.data.frame(remdat_complete)
  
  remdat_complete$delta.dist = remdat_complete$distance - 
    remdat_complete$non.distance 
  remdat_complete$delta.wkday = remdat_complete$weekday - 
    remdat_complete$non.weekday
  remdat_complete$delta.rec = remdat_complete$rec - 
    remdat_complete$non.rec
  remdat_complete$one = 1
  
  return(remdat_complete) 
  
}
```

```{r eval=FALSE}
n_sim <- 1
shifted_rems <- sapply(1:n_sim, function(x)
  remdat_shifting(remdats_nocox[[x]][[1]], remdats_nocox[[x]][[2]], 
                  p = p, n = n, seed=10), simplify=FALSE)
head(shifted_rems)
all((shifted_rems[[1]]$non.tms%%7>=2) ==shifted_rems[[1]]$non.weekday)
all(shifted_rems[[1]]$s == remdats_nocox[[1]][[1]]$s)
all(shifted_rems[[1]]$r == remdats_nocox[[1]][[1]]$r)
save(shifted_rems, file="output/running_example_shifted_data.RData")
```

```{r eval=FALSE}
gam.fits_shifted <- lapply(shifted_rems, function(x){
  dist_MAT <- cbind(as.numeric(x$distance), 
                    as.numeric(x$non.distance))
  ONE <- cbind(x$one, -x$one)
  fit <- gam(one ~ s(dist_MAT, by=ONE, k=11, bs="bs") + 
               delta.rec + delta.wkday - 1,
             family="binomial",
             data=x)
  return(fit)
}
)
save(gam.fits_shifted, file="output/running_example_gam.fits_shifted.RData")
```