---
title: "Running Example: Inference via Poisson / CoxPH / C-Logit / Logistic Regression"
author: "Boschi, M."
date: "2026-04-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synthetic Data: Academics moving in US states. 

## Preliminarities

More details about the preliminaries can be found in the document: `01-simulation-study-implementation`.

```{r pressure, echo=FALSE, message=FALSE, warning=FALSE}
source("script/packages.R")
```

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
source("script/states.R")
source("script/input_distance.R")
source("script/true_coefficients.R")
```

```{r}
n_sim = 1
```

```{r}
load(file="output/running_example_data.RData")
load(file="output/running_example_gam.fits.RData")
load(file="output/running_example_shifted_data.RData")
load(file="output/running_example_gam.fits_shifted.RData")
```

```{r}
rem_data <- remdats_cox[[1]][[1]]
cox_data <- remdats_cox[[1]][[3]]
sampled_data <- remdats_cox[[1]][[4]]
```

```{r}
(x1 <- which(rem_data$rec==1))
(x2 <- which(cox_data$tms==rem_data[x1,"tms"] & cox_data$event==1))
(x3 <- which(sampled_data$tms==rem_data[x1,"tms"] & sampled_data$event==1))
```

## Poisson Regression

```{r}
poisson_data <- cox_data
poisson_data$logtm  <- log(poisson_data$stop-poisson_data$start)
```

```{r}
poisson_display <- poisson_data[x2:(x2+7),c("s", "r", "start", "stop", 
                                                    "event", "rec", "logtm")]
poisson_display$start <- round(poisson_display$start, 2)
poisson_display$stop <- round(poisson_display$stop, 2)
poisson_display$logtm <- round(poisson_display$logtm, 2)
poisson_display <- rbind(rep("...", ncol(poisson_display)),
                        poisson_display, 
                        rep("...", ncol(poisson_display)))
colnames(poisson_display) <- c("sender", "receiver", 
                              "start", "stop", "event",
                              "x", "log_exposure")
rownames(poisson_display) <- NULL

poisson_display <- rbind(poisson_display, 
                          rep(" ", ncol(poisson_display)))
poisson_display
```

```{r}
ft <- flextable(poisson_display) %>%
  set_table_properties(layout = "fixed", width = 1) %>%
  width(j = c(7), width = 1.4) %>%      # Adjust columns 1 to 6
  fontsize(size = 10, part = "all")
save_as_image(ft, path="pictures/inference/poisson_display.png")
```


```{r}
poisson_data$x <- poisson_data$rec
poisson_fit <-  glm(event ~ x + offset(logtm), 
      family = poisson(link = "log"), 
      data = poisson_data)
summary(poisson_fit)
```

## CoxPH Regression

```{r}
cox_data_display <- cox_data[x2:(x2+7),c("s", "r", "start", "stop", 
                                          "event", "rec")]
cox_data_display$start <- round(cox_data_display$start, 2)
cox_data_display$stop <- round(cox_data_display$stop, 2)
cox_data_display <- rbind(rep("...", ncol(cox_data_display)),
                        cox_data_display, 
                        rep("...", ncol(cox_data_display)))
colnames(cox_data_display) <- c("sender", "receiver", 
                              "start", "stop", "event",
                              "x")
rownames(cox_data_display) <- NULL
cox_data_display <- rbind(cox_data_display, 
                          rep(" ", ncol(cox_data_display)))
cox_data_display
```

```{r}
ft <- flextable(cox_data_display) %>%
  set_table_properties(layout = "fixed", width = 1) %>%
  fontsize(size = 10, part = "all")
save_as_image(ft, path="pictures/inference/cox_data_display.png")
cox_data_display
```

```{r}
cox_data$x <- cox_data$rec
cox_fit <- coxph(Surv(start,stop,event) ~ x, data = cox_data)
summary(cox_fit)
```

```{r}
clogit_fit <- clogit(event ~ x + strata(tms), data = cox_data)
summary(clogit_fit)
```

## Conditional Logistic Regression in Practice 

```{r}
sampled_data_display <- sampled_data[x3:(x3+7),c("s", "r", "start", "stop", 
                                          "event", "rec")]
sampled_data_display$start <- round(sampled_data_display$start, 2)
sampled_data_display$stop <- round(sampled_data_display$stop, 2)
sampled_data_display <- rbind(rep("...", ncol(sampled_data_display)),
                        sampled_data_display, 
                        rep("...", ncol(sampled_data_display)))
colnames(sampled_data_display) <- c("sender", "receiver", 
                              "start", "stop", "event",
                              "x")
sampled_data_display <- rbind(sampled_data_display, 
                          rep(" ", ncol(sampled_data_display)))
rownames(sampled_data_display) <- NULL
sampled_data_display
```


```{r}
ft <- flextable(sampled_data_display) %>%
  set_table_properties(layout = "fixed", width = 1) %>%
  fontsize(size = 10, part = "all")
save_as_image(ft, path="pictures/inference/sampled_data_display.png")
```

```{r}
sampled_data$x <- sampled_data$rec
clogit_fit <- clogit(event ~ x + strata(tms), data = sampled_data)
summary(clogit_fit)
```

## Case-Control Partial Likelihood via GAM

```{r}
rem_data_display <- rem_data[x1:(x1+4),c("tms", "s", "r", "non.s", "non.r",
                                         "rec", "non.rec", "delta.rec")]
rem_data_display$tms <- round(rem_data_display$tms, 2)
rem_data_display$id_ev <- 1
rem_data_display$id_non_ev <- -1
rem_data_display$y <- 1
rem_data_display <- rbind(rep("...", ncol(rem_data_display)),
                        rem_data_display, 
                        rep("...", ncol(rem_data_display)))
rem_data_display <- rbind(rem_data_display, 
                          rep(" ", ncol(rem_data_display)))
colnames(rem_data_display) <- c("time", "sender_ev", "receiver_ev", 
                                "sender_non_ev", "receiver_non_ev", 
                                "x_ev", "x_non_ev", "delta_x", 
                                "id_ev", "id_non_ev", "y")
rownames(rem_data_display) <- NULL
rem_data_display
```

```{r}
ft <- flextable(rem_data_display) %>%
  set_table_properties(layout = "fixed", width = 1) %>%
  fontsize(size = 9, part = "all")
save_as_image(ft, path="pictures/inference/rem_data_display.png")
```

## Shifted Case-Control Partial Likelihood via GAM

```{r}
shifted_data <- shifted_rems[[1]][1:100,]
head(shifted_data)
```

```{r}
all(rem_data$s==shifted_data$s)
all(rem_data$r==shifted_data$r)
colnames(shifted_data)
```

```{r}
shifted_data_display <- shifted_data[x1:(x1+4),c("tms", "s", "r", 
                                                "non.tms", "non.s", "non.r",
                                                "weekday", "non.weekday", 
                                                "delta.wkday")]
shifted_data_display$tms <- round(shifted_data_display$tms, 2)
shifted_data_display$non.tms <- round(shifted_data_display$non.tms, 2)
shifted_data_display$y <- 1
shifted_data_display <- rbind(rep("...", ncol(shifted_data_display)),
                        shifted_data_display, 
                        rep("...", ncol(shifted_data_display)))
shifted_data_display <- rbind(shifted_data_display, 
                          rep(" ", ncol(shifted_data_display)))
colnames(shifted_data_display) <- c("time_ev", "sender_ev", "receiver_ev", 
                                    "time_non_ev", "sender_non_ev", 
                                    "receiver_non_ev", 
                                    "weekday_ev", "weekday_non_ev", 
                                    "delta_weekday", "y")
rownames(shifted_data_display) <- NULL
shifted_data_display
```

```{r}
ft <- flextable(shifted_data_display) %>%
  set_table_properties(layout = "fixed", width = 1) %>%
  fontsize(size = 9, part = "all")
save_as_image(ft, path="pictures/inference/shifted_data_display.png")
```

```{r}
shifted_ncc_data <- shifted_data
shifted_ncc_data$y <- 1
shifted_ncc_data$delta_weekday <- shifted_data$delta.wkday
gam(y ~ -1 + delta_weekday, family="binomial"(link = "logit"),
    data = shifted_ncc_data)
summary(gam(y ~ -1 + delta_weekday, family="binomial"(link = "logit"),
    data = shifted_ncc_data))
```

## Case-Control Partial Likelihood via GAM (estimating random effects)

```{r}
rem_data_display <- rem_data[x1:(x1+4),c("tms", "s", "r", "non.s", "non.r")]
rem_data_display$tms <- round(rem_data_display$tms, 2)
rem_data_display$id_ev <- 1
rem_data_display$id_non_ev <- -1
rem_data_display$y <- 1
ncc_data <- rem_data_display
colnames(rem_data_display) <- colnames(ncc_data) <- c("time", "sender_ev", 
                                                      "receiver_ev", 
                                "sender_non_ev", "receiver_non_ev", 
                                "id_ev", "id_non_ev", "y")
rem_data_display <- rbind(rep("...", ncol(rem_data_display)),
                        rem_data_display, 
                        rep("...", ncol(rem_data_display)))
rem_data_display <- rbind(rem_data_display, 
                          rep(" ", ncol(rem_data_display)))
rownames(rem_data_display) <- NULL
rem_data_display
```

```{r}
ft <- flextable(rem_data_display) %>%
  set_table_properties(layout = "fixed", width = 1) %>%
  fontsize(size = 9, part = "all")
save_as_image(ft, path="pictures/inference/rem_data_display_random_effect.png")
```

```{r}
by_matrix = cbind(ncc_data$id_ev,-ncc_data$id_ev)	
sender_act <- factor(c(ncc_data$sender_ev,ncc_data$sender_non_ev))
dim(sender_act) <- c(nrow(ncc_data), 2)
receiver_pop <- factor(c(ncc_data$receiver_ev,ncc_data$receiver_non_ev))
dim(receiver_pop) <- c(nrow(ncc_data), 2)
gam(y ~ s(sender_act, by=by_matrix, bs="re") + 
        s(receiver_pop, by=by_matrix, bs="re")  - 1, 
    family="binomial"(link = 'logit'),
    data=ncc_data)
```

```{r}
# rem_data <- remdats_nocox[[1]][[1]]
# rem_data_display <- rem_data[,c("tms", "s", "r", "non.s", "non.r",
#                                          "rec", "non.rec", "delta.rec")]
# rem_data_display$tms <- round(rem_data_display$tms, 2)
# rem_data_display$id_ev <- 1
# rem_data_display$id_non_ev <- -1
# rem_data_display$y <- 1
# colnames(rem_data_display) <- c("time", "sender_ev", "receiver_ev", 
#                                 "sender_non_ev", "receiver_non_ev", 
#                                 "x_ev", "x_non_ev", "delta_x", 
#                                 "id_ev", "id_non_ev", "y")
# rownames(rem_data_display) <- NULL
# ncc_data <- rem_data_display
# x_matrix <- cbind(ncc_data$x_ev, ncc_data$x_non_ev)
# by_matrix <- cbind(ncc_data$id_ev, ncc_data$id_non_ev)
# gam(y ~ -1 + s(x_matrix, by=by_matrix, k=4),
#         family="binomial"(link = "logit"), data = ncc_data)
```
