---
title: "Empirical Applications (1): World Trade Center"
author: "Martina Boschi"
date: "2026-04-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Basics: Modeling Phone Calls in an Emergency Context

## Preliminarities

The following section aims to load the necessary packages and import the data. Furthermore, we modify the data to facilitate their use within `eventnet`.

Specifically:
- We include rows that list the actors in the system. This allows us to construct the risk set.
- We provide the exogenous information in a format that can be read as actor-specific attributes by `eventnet`.

```{r pressure, echo=FALSE, message=FALSE, warning=FALSE}
source("script/packages.R")
```

```{r}
load(file="input/WTC.RData")
```

```{r}
data <- WTCPoliceCalls[,c("source", "recipient", "number")]
colnames(data) <- c("s1", "r1", "time")
data$type <- "call"
```

```{r}
num.actors <- length(WTCPoliceIsICR)
actors <- 1:num.actors
```

```{r}
before.min.time <- min(data$time) - 1
dummy.events <- data.frame(actors, actors,
                           rep(before.min.time,num.actors),
                           rep("add.actor",num.actors))
colnames(dummy.events) <- colnames(data)
head(dummy.events)
```

```{r}
events <- rbind(dummy.events, data)
```

```{r}
ICR <- which(WTCPoliceIsICR)
num.ICR <- length(ICR)
dummy.events <- data.frame(ICR, ICR,
                           rep(before.min.time,num.ICR),
                           rep("is.ICR",num.ICR))
colnames(dummy.events) <- colnames(events)
events <- rbind(dummy.events, events)
```

```{r}
events <- events[order(events$time),]
write.csv(events, file="input/WTC_data.csv", row.names = F)
```

The dataset `WTC_data.csv` is then processed in `eventnet`, and the corresponding configuration file can be found at `eventnet/configuration_WTC.txt`.

