---
title: "Empirical Applications (1): World Trade Center"
author: "Martina Boschi"
date: "2026-04-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Basics: Modeling Phone Calls in an Emergency Context

## Preliminarities

The following section aims to load the necessary packages and import the data. Furthermore, we modify the data to facilitate their use within `eventnet`.

Specifically:
- We include rows that list the actors in the system. This allows us to construct the risk set.
- We provide the exogenous information in a format that can be read as actor-specific attributes by `eventnet`.

```{r pressure, echo=FALSE, message=FALSE, warning=FALSE}
source("script/packages.R")
```

```{r eval=FALSE}
load(file="input/WTC.RData")
```

```{r eval=FALSE}
data <- WTCPoliceCalls[,c("source", "recipient", "number")]
colnames(data) <- c("s1", "r1", "time")
data$type <- "call"
```

```{r eval=FALSE}
num.actors <- length(WTCPoliceIsICR)
actors <- 1:num.actors
```

```{r eval=FALSE}
before.min.time <- min(data$time) - 1
dummy.events <- data.frame(actors, actors,
                           rep(before.min.time,num.actors),
                           rep("add.actor",num.actors))
colnames(dummy.events) <- colnames(data)
head(dummy.events)
```

```{r eval=FALSE}
events <- rbind(dummy.events, data)
```

```{r eval=FALSE}
ICR <- which(WTCPoliceIsICR)
num.ICR <- length(ICR)
dummy.events <- data.frame(ICR, ICR,
                           rep(before.min.time,num.ICR),
                           rep("is.ICR",num.ICR))
colnames(dummy.events) <- colnames(events)
events <- rbind(dummy.events, events)
```

```{r eval=FALSE}
events <- events[order(events$time),]
write.csv(events, file="input/WTC_data.csv", row.names = F)
```

The dataset `WTC_data.csv` is then processed in `eventnet`, and the corresponding configuration file can be found at `eventnet/configuration_WTC.txt`.

## Model fitting

```{r}
data_original <- read.csv("input/WTC_data_FULL.csv")
data_full <- data_original[,setdiff(colnames(data_original), 
                                    c("TYPE", 
                                      "EVENT_INTERVAL_ID", "EVENT", 
                                      "INTEGER_TIME", "TIME_POINT", 
                                      "TIME_UNIT", "num.actors"))]
rm(data_original)
head(data_full)
```

```{r}
clogit_fit_full <- clogit(IS_OBSERVED ~ 
                          + sender_activity
                          + receiver_popularity 
                          + repetition 
                          + reciprocity
                          + closure
                          + is.ICRs 
                          + is.ICRr
                          + strata(EVENT_INTERVAL)
                          , data = data_full)
s_clogit_fit_full <- summary(clogit_fit_full)
```

### Results

```{r}
s_clogit_fit_full
```


```{r}
names <- c("Sender Activity","Receiver Popularity","Repetition",
           "Reciprocity", "Closure", "Sender is ICR", "Receiver is ICR")

summary <- as.data.frame(s_clogit_fit_full$coefficients[,c(1:3)])
summary <- round(summary, 3)
summary <- cbind(rownames(s_clogit_fit_full$coefficients), summary)
colnames(summary) <- c("Explanatory Variable",
                       "Estimate", 
                       "Exp(Estimate)",
                       "S.E.(Estimate)")
rownames(summary) <- NULL
```

```{r}
ft <- flextable(summary) %>%
  set_table_properties(layout = "fixed", width = 1) %>%
  width(j = c(1), width = 2) %>%      
  fontsize(size = 10, part = "all")
save_as_image(ft, path="pictures/empirical/WTC/summary_WTC.png")
```

## Case-m-control sampling

```{r}
ensure_non_events <- function(data_sampled, data_full, m = 1, seed = 1234) {
  
  set.seed(seed)
  
  # Separate events and non-events from sampled data
  events_sampled <- data_sampled %>% filter(IS_OBSERVED == 1)
  non_events_sampled <- data_sampled %>% filter(IS_OBSERVED == 0)
  
  # Group events by TIME and EVENT_INTERVAL
  events_grouped <- events_sampled %>%
    distinct(TIME, EVENT_INTERVAL)
  
  # For each event group, check number of available non-events
  additions <- events_grouped %>%
    rowwise() %>%
    do({
      t <- .$TIME
      ei <- .$EVENT_INTERVAL
      
      # Existing non-events for this (TIME, EVENT_INTERVAL)
      current_non_events <- non_events_sampled %>%
        filter(TIME == t, EVENT_INTERVAL == ei)
      
      n_missing <- m - nrow(current_non_events)
      
      if (n_missing > 0) {
        # Get candidates from full data
        candidates <- data_full %>%
          filter(IS_OBSERVED == 0, TIME == t, EVENT_INTERVAL == ei)
        
        # Sample min(n_missing, number of available candidates)
        n_to_sample <- min(n_missing, nrow(candidates))
        
        if (n_to_sample > 0) {
          sampled <- candidates %>% sample_n(n_to_sample)
          sampled
        } else {
          tibble()  # no candidates available
        }
      } else {
        tibble()  # no missing non-events
      }
    }) %>%
    ungroup()
  
  # Combine original sampled data with additions
  data_fixed <- bind_rows(data_sampled, additions)
  
  return(data_fixed)
}
```

```{r}
sum(data_full$IS_OBSERVE==1)
nrow(data_full)/480
37*37 - 37
```

### m = 1

```{r}
data_original <- read.csv("input/WTC_data_NCC_1.csv")
data_sampled_1 <- data_original[,setdiff(colnames(data_original), 
                                    c("TYPE", "EVENT_INTERVAL_ID", "EVENT", 
                                      "INTEGER_TIME", "TIME_POINT", 
                                      "TIME_UNIT", "num.actors"))]
```

```{r}
sum(data_sampled_1$IS_OBSERVED==0)
sum(data_sampled_1$IS_OBSERVED==1)
```


```{r}
data_sampled_1 <- ensure_non_events(data_sampled_1, data_full, 
                                    m = 1, seed = 1234)
```

```{r}
sum(data_sampled_1$IS_OBSERVED==0)
sum(data_sampled_1$IS_OBSERVED==1)
```

```{r}
rm(data_original)
```

```{r}
clogit_fit_m1 <- clogit(IS_OBSERVED ~ 
                            + sender_activity
                          + receiver_popularity 
                          + repetition 
                          + reciprocity
                          + closure
                          + is.ICRs 
                          + is.ICRr
                          + strata(EVENT_INTERVAL)
                          , data = data_sampled_1)
s_clogit_fit_m1 <- summary(clogit_fit_m1)
s_clogit_fit_m1
```

### m = 5

```{r}
data_original <- read.csv("input/WTC_data_NCC_5.csv")
data_sampled_5 <- data_original[,setdiff(colnames(data_original), 
                                         c("TYPE", "EVENT_INTERVAL_ID", "EVENT", 
                                           "INTEGER_TIME", "TIME_POINT", 
                                           "TIME_UNIT", "num.actors"))]
rm(data_original)
```

```{r}
sum(data_sampled_5$IS_OBSERVED==0)
sum(data_sampled_5$IS_OBSERVED==1) * 5
```

```{r}
data_sampled_5 <- ensure_non_events(data_sampled_5, data_full, 
                                    m = 5, seed = 1)
```

```{r}
sum(data_sampled_5$IS_OBSERVED==0)
sum(data_sampled_5$IS_OBSERVED==1) * 5
```

```{r}
rm(data_original)
```

```{r}
clogit_fit_m5 <- clogit(IS_OBSERVED ~ 
                          + sender_activity
                        + receiver_popularity 
                        + repetition 
                        + reciprocity
                        + closure
                        + is.ICRs 
                        + is.ICRr
                        + strata(EVENT_INTERVAL)
                        , data = data_sampled_5)
s_clogit_fit_m5 <- summary(clogit_fit_m5)
s_clogit_fit_m5
```

### m = 20

```{r}
data_original <- read.csv("input/WTC_data_NCC_20.csv")
data_sampled_20 <- data_original[,setdiff(colnames(data_original), 
                                         c("TYPE", "EVENT_INTERVAL_ID", "EVENT", 
                                           "INTEGER_TIME", "TIME_POINT", 
                                           "TIME_UNIT", "num.actors"))]
```

```{r}
sum(data_sampled_20$IS_OBSERVED==0)
sum(data_sampled_20$IS_OBSERVED==1) * 20
```

```{r}
data_sampled_20 <- ensure_non_events(data_sampled_5, data_full, 
                                    m = 20, seed = 1234)
```

```{r}
sum(data_sampled_20$IS_OBSERVED==0)
sum(data_sampled_20$IS_OBSERVED==1) * 20
```

```{r}
clogit_fit_m20 <- clogit(IS_OBSERVED ~ 
                          + sender_activity
                        + receiver_popularity 
                        + repetition 
                        + reciprocity
                        + closure
                        + is.ICRs 
                        + is.ICRr
                        + strata(EVENT_INTERVAL)
                        , data = data_sampled_20)
s_clogit_fit_m20 <- summary(clogit_fit_m20)
s_clogit_fit_m20
```

### Results

```{r}
# Extract coefficients
coef_full <- coefficients(clogit_fit_full)
coef_m20  <- coefficients(clogit_fit_m20)
coef_m5   <- coefficients(clogit_fit_m5)
coef_m1   <- coefficients(clogit_fit_m1)

# Extract standard errors
se_full <- s_clogit_fit_full$coefficients[,3]
se_m20  <- s_clogit_fit_m20$coefficients[,3]
se_m5   <- s_clogit_fit_m5$coefficients[,3]
se_m1   <- s_clogit_fit_m1$coefficients[,3]

# Variable names
variables <- names(coef_full)

# Create table with formatted coefficient (SE)
results_table <- data.frame(
  Variable = variables,
  `Full Model` = sprintf("%.3f (%.3f)", coef_full, se_full),
  `m20 Model`  = sprintf("%.3f (%.3f)", coef_m20, se_m20),
  `m5 Model`   = sprintf("%.3f (%.3f)", coef_m5, se_m5),
  `m1 Model`   = sprintf("%.3f (%.3f)", coef_m1, se_m1),
  stringsAsFactors = FALSE
)
```

```{r}
latex_table <- xtable(results_table, 
                      caption = "Coefficients and Standard Errors from Conditional Logistic Regression Models", 
                      label = "tab:clogit_results", 
                      align = c("l", "l", "c", "c", "c", "c"))
print(latex_table, include.rownames = FALSE, 
      sanitize.text.function = identity)
```

